|

### 2023-02-05 manuscript/manuscript.Rmd
updating manuscript again

\begin{abstract}
Patient reported outcomes including quality of life assessments are increasingly being included as either primary or secondary outcomes in randomized controlled trials. While making the outcomes more relevant for patients it entails a challenge in cases where death or a similar event makes the outcome of interest undefined. A pragmatic - and much used - solution is to assign diseased patient with the lowest possible quality of life score. This makes medical sense, but creates a statistical problem since traditional tests such as the t-test or the Wilcoxon test potentially loses high amounts of statistical power. In this paper we propose a novel test that can keep the medically relevant composite outcome but preserve full statistical power. The test is also applicable in other situations where a specific value (say zero days alive outside hospitals) encodes a special meaning. The test is implemented in an R package. The paper includes a simulation study and an application to a COVID-19 randomized controlled trial.
\end{abstract}

\begin{center}
\textbf{Keywords:} Quality of Life, RCTs, Truncated Outcomes, Zero-inflation, 
\end{center}

@@

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), hence henceforth Wilcoxon, since distributions are rarely normal or symmetric. When all patients are alive and able and willing to provide quality of life scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of quality of life This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon and which is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare groups. This paper will demonstrate that this approach can lead to dramatic loss of statistical power. The paper will also present an alternative and novel statistical test that avoids the power loss. The novel test procedure is implemented in R and made publicly available.

The key-insight in the proposed test is to incorporate that the outcome is actually a two-dimensional outcome of a very special type, and that the constructed combined outcome follows a continuous-singular mixture distribution. This unusual distribution is why one cannot resort to non-parametric Wilcoxon tests since the singular component in the distribution of the combined outcome will get reduced to simple ties. It is noted that the handling of ties in standard statistical software varies and is opaque. However, the handling of ties is not the main reason why the Wilcoxon test suffers power loss. The main reason is that the null-hypothesis in these Wilcoxon type tests (stochastic domination) does not handle the empirical fact that treatments might influence mortality and quality of life differently.

@@

We consider random variables $(Y, A, R, X)$ where $Y$ is the continuous outcome, $A$ is a binary variable being equal to $1$ if $Y$ is observed and equal to $0$ if $Y$ is unobserved/undefined, $R$ is a binary treatment indicator, and $X$ is a $p$-dimensional vector of additional covariates. The objective is to focus on the bivariate distribution of $(Y \mid A = 1), A \mid R, X$ under the primary null-hypothesis of no treatment effect given by $(Y \mid A = 1), A \perp R \mid X$. Even though we model the outcome as a bivariate random variable, a combined outcome is often used in practice. The combined outcome is derived such that it is equal to $Y$ if $A = 1$ (and $Y$ is observed), and otherwise (when $A = 0$) it is equal to some predetermined value. The combined outcome can be written as
\begin{align}
  \widetilde{Y} =  Y \mathbbm{1}(A = 1) + \mathcal{E} \mathbbm{1}(A = 0)\label{eq:combinedOutcome}
\end{align}
where $\mathcal{E}$ is a fixed atom assigned as the outcome value when $Y$ is unobserved, and $\mathbbm{1}$ denotes the indicator function. The semi-continuous distribution of $\widetilde{Y}$ is therefore a probabilistic mixture of a singular distribution at $\mathcal{E}$ and a continuous distribution over the domain of the random variable $Y \mid A = 1$. Thus the statistical challenge can be rephrased as assert whether the treatment affects the distribution of $\widetilde{Y}$.

The conditional expected value of the combined outcome in Equation (\ref{eq:combinedOutcome}) is given by
\begin{align*}
  E[\widetilde{Y} \mid R, X] = E[Y \mid R, X, A=1]P(A = 1 \mid R, X) + \mathcal{E}P(A = 0 \mid R, X)
\end{align*}
From this expression one can show that a treatment comparison expressed in terms of a contrast of the expectation of $\widetilde{Y}$ can be zero even though the distribution of $((Y \mid A=1), A)$  depends on $R$. To see this, let $E[Y \mid A=1, R = 0, X] = \mu(X)$ and $E[Y \mid A=1, R = 1, X] = \mu(X) + \mu_{\delta}(X)$ and assume without loss of generalization that $\mathcal{E} = 0$. Then the average treatment effect for the combined outcome conditional on baseline covariates is 
\begin{align}
  \Delta(X) &= E[\widetilde{Y} \mid R = 1, X] - E[\widetilde{Y} \mid R = 0, X]\label{eq:Delta}\\
        &= \mu(X)\left[P(A = 1 \mid R=1,X) - P(A = 1 \mid R = 0, X)\right] + \mu_{\delta}(X) P(A = 1 \mid R = 1, X)\nonumber
\end{align}
If we auspiciously let $\mu_{\delta}(X) = \mu(X)\left[P(A = 1 \mid R = 0, X) P(A = 1 \mid R = 1, X)^{-1} - 1\right]$ then $\Delta(X) = 0$ for all values of $X$ even though $\mu_{\delta}(X) \ne 0$ and $P(A = 1 \mid R = 0, X) \ne P(A = 1 \mid R = 1, X)$. On the other hand, if $\mu_\delta(X) = 0$ and $P(A = 1 \mid R=1,X) = P(A = 1 \mid R = 0, X)$ then $\Delta(X)$ is necessarily equal to zero. This illustrates that a significance test of no treatment effect must have two degrees of freedom. Such a test which we develop in the subsequent sections is conceptually difference than a test for the null-hypothesis $\Delta(X) = 0$. Noe that in RCTs one would typically not include any variables $X$ since these are balanced by design.

We propose to test the null-hypothesis of no treatment effect on the combined outcome by a likelihood ratio test of the joint distribution of $Y_i \mid A_i$ and $A_i$. This has the advantage that it increases the efficiency compared the Wilcoxon test and it yields a single p-value appropriate for testing a primary outcome in a clinical trial. Let $(Y_i, A_i, R_i, X_i)$ for $i = 1, \ldots, n$ be independent and identically distributed random variables. We can write our model in a general form as 
\begin{align*}
  Y_i \mid A_i = 1, R_i, X_i &\sim F(Y_i; \mu_i(R_i, X_i), \Psi)\\
  A_i \mid R_i, X_i &\sim \text{Bernoulli}(A_i; \pi_i(R_i, X_i))  
\end{align*}
for some mean functions $\pi_i$ and $\mu_i$ and a distribution function $F$ characterizing the distribution of the observed continuous outcomes with possible nuisance parameter $\Psi$. The joint likelihood function for the combined outcome conditional on treatment and baseline covariates is
\begin{align*}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i), \Psi) &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}P(A_i = a_i; \pi_i(R_i, X_i))\\
     &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i} \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}\nonumber
\end{align*}
We note that when $F$ admits an absolutely continuous density function the likelihood function can equivalently be written solely in terms of the combined outcome $\widetilde{Y}$ in Equation (\ref{eq:combinedOutcome}) since $f(Y_i; \cdot)^{A_i} = f(\widetilde{Y}_i; \cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ and $\pi_i(\cdot)^{A_i} = \pi_i(\cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ almost surely. An important property of this model is that the likelihood function then factorizes into two components as
\begin{align}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i)) &= L_{1,n}(\mu_i(R_i, X_i), \Psi) L_{2,n}(\pi_i(R_i, X_i))\label{eq:likFactorization}
\end{align}
where $L_{1,n}(\mu_i(R_i, X_i), \Psi) = \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}$ and $L_{2,n}(\pi_i(R_i, X_i)) = \prod_{i=1}^n \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}$. This implies that the parameters for the observed outcomes, $\mu_i(R_i, X_i)$, can be estimated independently of the parameters governing the probability of observing the outcome, $\pi_i(R_i, X_i)$. This factorization is a consequence of the likelihood construction and it does neither assume nor require independence between the value of the outcome and the probability of observing it.

@@

 use the profile likelihood ratio test statistic which can be written as a function of the treatment effects as
and the value of the likelihood ratio test statistic under the null-hypothesis of no treatment effect is therefore $W_n(0, 0)$. Similar to the factorization of the likelihood function in Equation (\ref{eq:likFactorization}), the likelihood ratio test statistic in Equation (\ref{eq:lrtgeneral})  also decomposes into the sum
\begin{align}
  W_n(\mu_\delta, \beta_\delta) &= W_{1,n}(\mu_\delta) + W_{2,n}(\beta_\delta)\label{eq:LRTsum} 
\end{align}
of two likelihood ratio test statistics -- one for the continuous part and one for the discrete part of the combined outcome. Under very general conditions it follows that $W_n(\mu_\delta, \beta_\delta)$ is approximately $\chi^2$ distributed with two degrees of freedom [@wilks1938large; @owen1988empirical].

@@

With the generalized linear models in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}) the first term is the LRT statistic with a single degree of freedom in a linear regression model, and the second term is the LRT statistic with a single degree of freedom in a logistic regression model. This makes this test very easy to perform in standard statistical software that outputs the likelihood value for a model fit without the need of special methods. The calculation of the likelihood ratio test simply amounts to estimating two linear regression models and two logistic regression models -- for each type a model with and a model without the binary treatment indicator as a covariate. Combining the two likelihood ratios according to Equation (\ref{eq:LRTsum}) calculates our test statistic, and the p-value for no treatment effect can be determined through the $\chi^2$-distribution with two degrees of freedom. The critical value for rejecting the null-hypothesis of no treatment effect at the 5% level is equal to $5.99$ and equal to $9.21$ at the 1% level.

@@

\begin{proposition}
Assume that $P(A_i = 1 \mid R_i = r) > 0$ for $r = 0,1$ and let $m_j = \sum_{i=1}^n 1_{R_i = j}$. Then the parametric profile likelihood ratio test statistic $W_n^P(0,0)$ stated in Equation (\ref{eq:LRTtotal}) for the null-hypothesis of no treatment effect on the combined outcome is asymptotically $\chi^2$ distributed with two degrees of freedom for $m_1, m_2 \rightarrow \infty$ and $m_1/m_2 \rightarrow c$ for some finite, positive constant $c$.
\end{proposition}

@@

Let $\mathcal{I}_0 = \left\{i = 1,\ldots, n : R_i = 0, A_i = 1\right\}$ and $\mathcal{I}_1 = \left\{i = 1,\ldots, n : R_i = 1, A_i = 1\right\}$ be the sets of indices of the observed outcomes for the two treatments. The empirical likelihood ratio test statistic as a function of the difference in expected value is given by
\begin{align*}
  W_{1,n}^E(\mu_\delta^\ast) &= 2\sup_{\mu}\left(\sum_{i \in \mathcal{I}_0}\log \left(1 + \lambda_1 (Y_i - \mu)\right) + \sum_{j \in \mathcal{I}_1}\log\left(1 + \lambda_2 (Y_j - \mu - \mu_\delta^\ast)\right)\right)
\end{align*}
where $\lambda_1$ and $\lambda_2$ are the solutions to the following equations
\begin{align*}
  |\mathcal{I}_0|^{-1}\sum_{i \in \mathcal{I}_0} \frac{Y_i - \mu}{1 + \lambda_1 (Y_i - \mu)} = 0, \quad |\mathcal{I}_1|^{-1}\sum_{j \in \mathcal{I}_1}\frac{Y_j - \mu - \mu_\delta^\ast}{1 + \lambda_2 (Y_j - \mu - \beta_\delta^\ast)} = 0
\end{align*}
We refer to the appendix for a derivation of the test statistics.

The semi-parametric likelihood ratio test statistic for testing the null-hypothesis of no treatment effect is equal to $W_n^{SP}(0, 0)$ where 
\begin{align}
  W_n^{SP}(\mu_\delta^\ast, \alpha_\delta^\ast) = W_{1,n}^E(\mu_\delta^\ast) + W_{2,n}(\alpha_\delta^\ast) \label{eq:LRT_SP_total}
\end{align}
By the non-parametric Wilk's theorem [@owen1988empirical] it follows that $W_n^{SP}(0, 0) \sim \chi^2_2$ asymptotically.

\begin{proposition}
Assume that $P(A_i = 1 \mid R_i = r) > 0$ for $r = 0,1$ and let $m_j = \sum_{i=1}^n 1_{R_i = j}$. Then the semi-parametric profile likelihood ratio test statistic $W_n^{SP}(0,0)$ stated in Equation (\ref{eq:LRT_SP_total}) for the null-hypothesis of no treatment effect on the combined outcome is asymptotically $\chi^2$ distributed with two degrees of freedom for $m_1, m_2 \rightarrow \infty$ and $m_1/m_2 \rightarrow c$ for some finite, positive constant $c$.
\end{proposition}

@@

The observed difference in means for the combined outcome, $\Delta$ in Equation (\ref{eq:Delta}), is `r round(mean(TruncCompExample$Y[TruncCompExample$R == 1])  - mean(TruncCompExample$Y[TruncCompExample$R == 0]), 3)` and both a two-sample t-test and a Wilcoxon rank sum test show highly insignificant effects of the treatment with p-values of `r round(t.test(Y ~ R, data = TruncCompExample)$p.value, 3)` and `r round(wilcox.test(Y ~ R, data = TruncCompExample)$p.value, 3)` respectively. In order to analyse the data using proposed method we use the function \texttt{truncComp} as follows
```{r}
model <- truncComp(Y ~ R, atom = 0, data = TruncCompExample, method = "SPLRT")
```
where the formula interface is similar to other regression models implemented in R. The argument \texttt{atom} identifies the value assigned to the truncated outcomes, and \texttt{method} can be \texttt{SPLRT} or \texttt{LRT} for either the semi-parametric or the parametric likelihood ratio test respectively. In this example we have opted for the semi-parametric version to show how easily this additional flexibility is included. We obtain the results of the estimation by a calling the function \texttt{summary} on the model:
```{r, comment=''}
summary(model)
```
The output from the call to \texttt{summary} displays the estimates for the two treatment contrasts corresponding to $\mu_\delta$ and $\exp(\alpha_\delta)$ in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}). These contrasts quantify the difference in means among the observed outcomes and the odds ratio of being observed respectively in accordance with the model specification. Each estimated treatment contrast is accompanied by a 95\% confidence interval, and finally the output displays the joint likelihood ratio test statistic and the associated $p$-value for the joint null-hypothesis of no treatment effect.

From the output we see that the semi-parametric likelihood ratio analysis reports an extremely low $p$-value for null-hypothesis of no joint treatment effect. This strongly contradicts the conclusions from both the previous t-test and Wilcoxon test. The confidence intervals for the two treatment contrasts indicate that the average value for the observed outcome in the group defined by $R = 1$ is significantly higher than the average value for the observed outcome in the group with $R = 0$.

@@

To assess the power we vary the sample size between 50 and 350 observations in each group with increments of 25, and for each scenario we perform 25,000 simulations and estimate the power as the proportion of rejected null hypotheses at the 5\% level. We compare the power of both our parametric and semi-parametric likelihood ratio tests to the two-sample t-test and the Wilcoxon test. Figure \ref{fig:powerCurves} shows the estimated power functions. The results are also presented in table form in the supplementary material.

@@

In addition to the power simulations we also performed a simulation study for the type I error. We considered two scenarios similar to scenarios 1 and 4 in Table \ref{tab:simulation} under a null-hypothesis of no treatment effect. Specifically, we considered $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim N(3,1)$ and $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim \Beta(1, 1)$. We estimated the type I error by simulations for different sample sizes and $\pi \in \left\{0.2, 0.4, 0.6, 0.8\right\}$. The results are presented in the supplementary material.

@@

Comparing the groups by a Wilcoxon test yields a p-value of $0.242$ hence leading to a non-significant conclusion at the pre-specified 5\% level. Using our proposed parametric likelihood ratio test yields a p-value of $0.019$ and therefore leading to a rejection of the null-hypothesis. From the histograms it could be questioned whether the continuous part of the distributions are normal distributed. Accordingly, we also apply our proposed semi-parametric version of the likelihood ratio test. This yields a p-value of $0.021$. In conclusion, our proposed method is able to detect the difference between the groups. Further and in contrast to the Wilcoxon test, our method also provides a way to interpret the difference between the groups. This is illustrated in the right panel of Figure \ref{fig:application} showing simultaneous confidence regions for the two effect parameters. In this case it is clear that the primary effect of the intervention is not on the discrete component (i.e., the proportion of patients who never get off life support). Instead, the effect appears to be on the continuous part and is between half a day and seven days.

>>>>

\begin{abstract}
Patient reported outcomes including quality of life assessments are increasingly being included as either primary or secondary outcomes in randomized controlled trials. While making the outcomes more relevant for patients it entails a challenge in cases where death or a similar event makes the outcome of interest undefined. A pragmatic -- and much used -- solution is to assign diseased patient with the lowest possible quality of life score. This makes medical sense, but creates a statistical problem since traditional tests such as the t-test or the Wilcoxon test potentially loses high amounts of statistical power. In this paper we propose a novel test that can keep the medically relevant composite outcome but preserve full statistical power. The test is also applicable in other situations where a specific value (say zero days alive outside hospitals) encodes a special meaning. The test is implemented in an R package, and the paper includes a simulation study and an application to a COVID-19 randomized controlled trial.
\end{abstract}

\begin{center}
\textbf{Keywords:} Quality of Life, Randomized Controlled Trials,\\ Truncated Outcomes, Semi-Continuous Outcomes, Empirical Likelihood
\end{center}

@@

In medical intervention research including in particular randomized controlled trials (RCTs) there is a trend towards increased use of patient reported outcomes; Quality of Life scores being one of the most prominent of these. Established statistical practice is to compare these between treatment groups using non-parametric tests such as Mann–Whitney–Wilcoxon rank-sum test (@wilcoxon3001968individual, @mann1947test), henceforth Wilcoxon, since distributions are rarely normal or symmetric. When all patients are alive and able and willing to provide quality of life scores at the scheduled measurement time this procedure works very well. Not all clinical settings, however, will have all participants alive at time of scheduled assessment of quality of life This is particularly true in trials within intensive care where mortality of approximately 30% is not uncommon and which is the setting that motivated this work. A pragmatic solution which is gaining popularity is to simply give the deceased patients the lowest possible score and then use a Wilcoxon test or similar to compare groups. This paper will demonstrate that this approach can lead to dramatic loss of statistical power. The paper will also present an alternative and novel statistical test that avoids the power loss. The novel test procedure is implemented in R and made publicly available.

The key-insight in the proposed test is to incorporate that the outcome is actually a two-dimensional outcome of a very special type, and that the constructed combined outcome follows a continuous-singular mixture distribution. This unusual distribution is why one cannot resort to non-parametric Wilcoxon tests since the singular component in the distribution of the combined outcome will get reduced to simple ties. It is noted that the handling of ties in standard statistical software varies and is opaque. However, the handling of ties is not the main reason why the Wilcoxon test suffers power loss. The main reason is that the null hypothesis in these Wilcoxon type tests (stochastic domination) does not handle the empirical fact that treatments might influence mortality and quality of life differently.

@@

We consider random variables $(Y, A, R, X)$ where $Y$ is the continuous outcome, $A$ is a binary variable being equal to $1$ if $Y$ is observed and equal to $0$ if $Y$ is unobserved/undefined, $R$ is a binary treatment indicator, and $X$ is a $p$-dimensional vector of additional covariates. The objective is to focus on the bivariate distribution of $(Y \mid A = 1), A \mid R, X$ under the primary null hypothesis of no treatment effect given by $(Y \mid A = 1), A \perp R \mid X$. Even though we model the outcome as a bivariate random variable, a combined outcome is often used in practice. The combined outcome is derived such that it is equal to $Y$ if $A = 1$ (and $Y$ is observed), and otherwise (when $A = 0$) it is equal to some predetermined value. The combined outcome can be written as
\begin{align}
  \widetilde{Y} =  Y \mathbbm{1}(A = 1) + \mathcal{E} \mathbbm{1}(A = 0)\label{eq:combinedOutcome}
\end{align}
where $\mathcal{E}$ is a fixed atom assigned as the outcome value when $Y$ is unobserved, and $\mathbbm{1}$ denotes the indicator function. The semi-continuous distribution of $\widetilde{Y}$ is therefore a probabilistic mixture of a singular distribution at $\mathcal{E}$ and a continuous distribution over the domain of the random variable $Y \mid A = 1$. Thus the statistical challenge can be rephrased as assert whether the treatment affects the distribution of $\widetilde{Y}$.

The conditional expected value of the combined outcome in Equation (\ref{eq:combinedOutcome}) is given by
\begin{align*}
  E[\widetilde{Y} \mid R, X] = E[Y \mid R, X, A=1]P(A = 1 \mid R, X) + \mathcal{E}P(A = 0 \mid R, X)
\end{align*}
From this expression one can show that a treatment comparison expressed in terms of a contrast of the expectation of $\widetilde{Y}$ can be zero even though the distribution of $((Y \mid A=1), A)$  depends on $R$. To see this, let $E[Y \mid A=1, R = 0, X] = \mu(X)$ and $E[Y \mid A=1, R = 1, X] = \mu(X) + \mu_{\delta}(X)$ and assume without loss of generalization that $\mathcal{E} = 0$. Then the average treatment effect for the combined outcome conditional on baseline covariates is 
\begin{align}
  \Delta(X) &= E[\widetilde{Y} \mid R = 1, X] - E[\widetilde{Y} \mid R = 0, X]\label{eq:Delta}\\
        &= \mu(X)\left[P(A = 1 \mid R=1,X) - P(A = 1 \mid R = 0, X)\right] + \mu_{\delta}(X) P(A = 1 \mid R = 1, X)\nonumber
\end{align}
If we auspiciously let $\mu_{\delta}(X) = \mu(X)\left[P(A = 1 \mid R = 0, X) P(A = 1 \mid R = 1, X)^{-1} - 1\right]$ then $\Delta(X) = 0$ for all values of $X$ even though $\mu_{\delta}(X) \ne 0$ and $P(A = 1 \mid R = 0, X) \ne P(A = 1 \mid R = 1, X)$. On the other hand, if $\mu_\delta(X) = 0$ and $P(A = 1 \mid R=1,X) = P(A = 1 \mid R = 0, X)$ then $\Delta(X)$ is necessarily equal to zero. This illustrates that a significance test of no treatment effect must have two degrees of freedom. Such a test which we develop in the subsequent sections is conceptually difference than a test for the null hypothesis $\Delta(X) = 0$. Noe that in RCTs one would typically not include any variables $X$ since these are balanced by design.

We propose to test the null hypothesis of no treatment effect on the combined outcome by a likelihood ratio test of the joint distribution of $Y_i \mid A_i$ and $A_i$. This has the advantage that it increases the efficiency compared the Wilcoxon test and it yields a single p-value appropriate for testing a primary outcome in a clinical trial. Let $(Y_i, A_i, R_i, X_i)$ for $i = 1, \ldots, n$ be independent and identically distributed random variables. We can write our model in a general form as 
\begin{align*}
  Y_i \mid A_i = 1, R_i, X_i &\sim F(Y_i; \mu_i(R_i, X_i), \Psi)\\
  A_i \mid R_i, X_i &\sim \text{Bernoulli}(A_i; \pi_i(R_i, X_i))  
\end{align*}
for some mean functions $\pi_i$ and $\mu_i$ and a distribution function $F$ characterizing the distribution of the observed continuous outcomes with possible nuisance parameter $\Psi$. The joint likelihood function for the combined outcome conditional on treatment and baseline covariates is
\begin{align*}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i), \Psi) &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}P(A_i = a_i; \pi_i(R_i, X_i))\\
     &= \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i} \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}\nonumber
\end{align*}
We note that when $F$ admits an absolutely continuous density function the likelihood function can equivalently be written solely in terms of the combined outcome $\widetilde{Y}$ in Equation (\ref{eq:combinedOutcome}) since $f(Y_i; \cdot)^{A_i} = f(\widetilde{Y}_i; \cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ and $\pi_i(\cdot)^{A_i} = \pi_i(\cdot)^{\mathbbm{1}(\widetilde{Y}_i \ne \mathcal{E})}$ almost surely. An important property of this model is that the likelihood function then factorizes into two components as
\begin{align}
 L_n(\mu_i(R_i, X_i), \pi_i(R_i, X_i)) &= L_{1,n}(\mu_i(R_i, X_i), \Psi) L_{2,n}(\pi_i(R_i, X_i))\label{eq:likFactorization}
\end{align}
where $L_{1,n}(\mu_i(R_i, X_i), \Psi) = \prod_{i=1}^n f(Y_i; \mu_i(R_i, X_i), \Psi)^{A_i}$ and $L_{2,n}(\pi_i(R_i, X_i)) = \prod_{i=1}^n \pi_i(R_i, X_i)^{A_i} (1 - \pi_i(R_i, X_i))^{1-A_i}$. This implies that the parameters for the observed outcomes, $\mu_i(R_i, X_i)$, can be estimated independently of the parameters governing the probability of observing the outcome, $\pi_i(R_i, X_i)$. This factorization is a consequence of the likelihood construction and it does neither assume nor require independence between the value of the outcome and the probability of observing it.

@@

 use the profile likelihood ratio test statistic which can be written as a function of the treatment effects as
and the value of the likelihood ratio test statistic under the null hypothesis of no treatment effect is therefore $W_n(0, 0)$. Similar to the factorization of the likelihood function in Equation (\ref{eq:likFactorization}), the likelihood ratio test statistic in Equation (\ref{eq:lrtgeneral})  also decomposes into the sum
\begin{align}
  W_n(\mu_\delta, \beta_\delta) &= W_{1,n}(\mu_\delta) + W_{2,n}(\beta_\delta)\label{eq:LRTsum} 
\end{align}
of two likelihood ratio test statistics -- one for the continuous part and one for the discrete part of the combined outcome. Under very general conditions it follows that $W_n(\mu_\delta, \beta_\delta)$ is approximately $\chi^2$ distributed with two degrees of freedom [@wilks1938large; @owen1988empirical].

@@

With the generalized linear models in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}) the first term is the LRT statistic with a single degree of freedom in a linear regression model, and the second term is the LRT statistic with a single degree of freedom in a logistic regression model. This makes this test very easy to perform in standard statistical software that outputs the likelihood value for a model fit without the need of special methods. The calculation of the likelihood ratio test simply amounts to estimating two linear regression models and two logistic regression models -- for each type a model with and a model without the binary treatment indicator as a covariate. Combining the two likelihood ratios according to Equation (\ref{eq:LRTsum}) calculates our test statistic, and the p-value for no treatment effect can be determined through the $\chi^2$-distribution with two degrees of freedom. The critical value for rejecting the null hypothesis of no treatment effect at the 5% level is equal to $5.99$ and equal to $9.21$ at the 1% level.

@@

\begin{proposition}
Assume that $P(A_i = 1 \mid R_i = r) > 0$ for $r = 0,1$ and let $m_j = \sum_{i=1}^n 1_{R_i = j}$. Then the parametric profile likelihood ratio test statistic $W_n^P(0,0)$ stated in Equation (\ref{eq:LRTtotal}) for the null hypothesis of no treatment effect on the combined outcome is asymptotically $\chi^2$ distributed with two degrees of freedom for $m_1, m_2 \rightarrow \infty$ and $m_1/m_2 \rightarrow c$ for some finite, positive constant $c$.
\end{proposition}

@@

Let $\mathcal{I}_0 = \left\{i = 1,\ldots, n : R_i = 0, A_i = 1\right\}$ and $\mathcal{I}_1 = \left\{i = 1,\ldots, n : R_i = 1, A_i = 1\right\}$ be the sets of indices of the observed outcomes for the two treatments. The empirical likelihood ratio test statistic as a function of the difference in expected value is given by
\begin{align*}
  W_{1,n}^E(\mu_\delta^\ast) &= 2\sup_{\mu}\left(\sum_{i \in \mathcal{I}_0}\log \left(1 + \lambda_1 (Y_i - \mu)\right) + \sum_{j \in \mathcal{I}_1}\log\left(1 + \lambda_2 (Y_j - \mu - \mu_\delta^\ast)\right)\right)
\end{align*}
where $\lambda_1$ and $\lambda_2$ are the solutions to the following equations
\begin{align*}
  |\mathcal{I}_0|^{-1}\sum_{i \in \mathcal{I}_0} \frac{Y_i - \mu}{1 + \lambda_1 (Y_i - \mu)} = 0, \quad |\mathcal{I}_1|^{-1}\sum_{j \in \mathcal{I}_1}\frac{Y_j - \mu - \mu_\delta^\ast}{1 + \lambda_2 (Y_j - \mu - \beta_\delta^\ast)} = 0
\end{align*}
We refer to the supplementary material for a derivation of this test statistics.

The semi-parametric likelihood ratio test statistic for testing the null hypothesis of no treatment effect is equal to $W_n^{SP}(0, 0)$ where 
\begin{align}
  W_n^{SP}(\mu_\delta^\ast, \alpha_\delta^\ast) = W_{1,n}^E(\mu_\delta^\ast) + W_{2,n}(\alpha_\delta^\ast) \label{eq:LRT_SP_total}
\end{align}
By the non-parametric Wilk's theorem [@owen1988empirical] it follows that $W_n^{SP}(0, 0) \sim \chi^2_2$ asymptotically.

\begin{proposition}
Assume that $P(A_i = 1 \mid R_i = r) > 0$ for $r = 0,1$ and let $m_j = \sum_{i=1}^n 1_{R_i = j}$. Then the semi-parametric profile likelihood ratio test statistic $W_n^{SP}(0,0)$ stated in Equation (\ref{eq:LRT_SP_total}) for the null hypothesis of no treatment effect on the combined outcome is asymptotically $\chi^2$ distributed with two degrees of freedom for $m_1, m_2 \rightarrow \infty$ and $m_1/m_2 \rightarrow c$ for some finite, positive constant $c$.
\end{proposition}

@@

The observed difference in means for the combined outcome, $\Delta$ in Equation (\ref{eq:Delta}), is `r round(mean(TruncCompExample$Y[TruncCompExample$R == 1])  - mean(TruncCompExample$Y[TruncCompExample$R == 0]), 3)` and both a two-sample t-test and a Wilcoxon rank sum test show highly insignificant effects of the treatment with p-values of `r round(t.test(Y ~ R, data = TruncCompExample)$p.value, 3)` and `r round(wilcox.test(Y ~ R, data = TruncCompExample)$p.value, 3)` respectively. In order to analyse the data using proposed method we use the function \texttt{truncComp} as follows
```{r}
model <- truncComp(Y ~ R, atom = 0, data = TruncCompExample, method = "SPLRT")
```
where the formula interface is similar to other regression models implemented in R. The argument \texttt{atom} identifies the value assigned to the truncated outcomes, and \texttt{method} can be \texttt{SPLRT} or \texttt{LRT} for either the semi-parametric or the parametric likelihood ratio test respectively. In this example we have opted for the semi-parametric version to show how easily this additional flexibility is included. We obtain the results of the estimation by a calling the function \texttt{summary} on the model:
```{r, comment=''}
summary(model)
```
The output from the call to \texttt{summary} displays the estimates for the two treatment contrasts corresponding to $\mu_\delta$ and $\exp(\alpha_\delta)$ in Equations (\ref{eq:glm1}) and (\ref{eq:glm2}). These contrasts quantify the difference in means among the observed outcomes and the odds ratio of being observed respectively in accordance with the model specification. Each estimated treatment contrast is accompanied by a 95\% confidence interval, and finally the output displays the joint likelihood ratio test statistic and the associated $p$-value for the joint null hypothesis of no treatment effect.

From the output we see that the semi-parametric likelihood ratio analysis reports an extremely low $p$-value for null hypothesis of no joint treatment effect. This strongly contradicts the conclusions from both the previous t-test and Wilcoxon test. The confidence intervals for the two treatment contrasts indicate that the average value for the observed outcome in the group defined by $R = 1$ is significantly higher than the average value for the observed outcome in the group with $R = 0$.

@@

To assess the power we vary the sample size between 50 and 350 observations in each group with increments of 25, and for each scenario we perform 25,000 Monte Carlo simulations and estimate the power as the proportion of rejected null hypotheses at the 5\% level. We compare the estimated power for both our parametric and semi-parametric likelihood ratio tests to the two-sample t-test and the Wilcoxon test. Figure \ref{fig:powerCurves} shows the estimated power functions. The results are also presented in table form in the supplementary material.

@@

In addition to the power simulations we also performed a simulation study for the type I error. We considered two scenarios similar to scenarios 1 and 4 in Table \ref{tab:simulation} under a null hypothesis of no treatment effect. Specifically, we considered $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim N(3,1)$ and $A_i \sim \Bernoulli(\pi)$, $Y_i \mid A_i = 1 \sim \Beta(1, 1)$. We estimated the type I error through Monte Carlo simulation for different sample sizes and $\pi \in \left\{0.2, 0.4, 0.6, 0.8\right\}$. The results are presented in the supplementary material.

@@

Comparing the groups by a Wilcoxon test yields a p-value of $0.242$ hence leading to a non-significant conclusion at the pre-specified 5\% level. Using our proposed parametric likelihood ratio test yields a p-value of $0.019$ and therefore leading to a rejection of the null hypothesis. From the histograms it could be questioned whether the continuous part of the distributions are normal distributed. Accordingly, we also apply our proposed semi-parametric version of the likelihood ratio test. This yields a p-value of $0.021$. In conclusion, our proposed method is able to detect the difference between the groups. Further and in contrast to the Wilcoxon test, our method also provides a way to interpret the difference between the groups. This is illustrated in the right panel of Figure \ref{fig:application} showing simultaneous confidence regions for the two effect parameters. In this case it is clear that the primary effect of the intervention is not on the discrete component (i.e., the proportion of patients who never get off life support). Instead, the effect appears to be on the continuous part and is between half a day and seven days.
